--- 
title: "Artifex Spring 2022"
author: "David Reynolds"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
# output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: openscapes/series
description: "Class materials for Paul 520."
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(kableExtra)
library(tidyverse)
library(lubridate)
library(openintro)
library(pander)
library(broom)
library(astsa)
```


# Welcome {#welcome}

Welcome to Artifex! The primary objective of this course/ club is to equip its members with the essential skills of a data scientist. The course delivery is a mix of lectures and project-based learning. The lectures will alternate between theory and tools. For the first half of the semester, the theory part of the course will be focused on linear regression and the tools part of the course will be related to R, R Studio, and Github. In the second half of the course, the theory will be focused on time series analysis and students will work in teams on a complex time series dataset hosted in Kaggle. 

Below, I have outlined the structure for the first half of this course. This is open to potential revision depending on whether students would like to go into more or less depth on any of the listed topics.



```{r}

t = mdy("1/24/2022")

Week_of = c(t, t + (1:7 * 7))

Topics = c("Regression I (Estimation)",
  "R fundamentals - functions",
  "Regression II (Inference)",
  "Github and R",
  "Regression III (Prediction)",
  "Exploratory Data Analysis",
  "Regression IV (Diagnostics)",
  "Spring Break!")

syll = data.frame("Week" = Week_of, Topic = Topics, Deliverables = c("Exercises 1", "Exercises 2", "Exercises 3", rep("TBD", 4), "Nada")) 

kbl(syll) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

**Helpful online resources**

- [Download R](https://cloud.r-project.org)
- [Download R Studio Desktop](https://www.rstudio.com/products/rstudio/download/)
- [R for Data Science](https://r4ds.had.co.nz/)

**Class Participation**

The vast majority of our work for this class will occur during our meetings. Therefore, attendance and participation is required by all students. Please email me at least 24 hours in advance if you have an excused absence. I also ask that if you are working within a project team that you contact your team about any planned or unplanned absence so they are not left at a disadvantage, can reassign work for if needed, or reconnect with you on any role assignments to complete.


